---
title: "UN data analysis"
date: "2023-08-22"
output: 
  html_document:
    code_folding: hide
    df_print: paged
    fig_width: 7
    fig_height: 6
    theme: cosmo
    highlight: zenburn
    toc: true
    number_sections: true
    toc_depth: 4
    toc_float:
      collapsed: false
  pdf_document:
    toc: true
    toc_depth: '4'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

The **UN** data set available in R is used for this analysis. 
It provides information on national health, welfare and population statistics for 213 locations, mainly UN members.
Different hypothesis are tested to gain insights into the differences in the groups and regions.


**fertility**: total fertility rate, number of children per woman

**ppgdp**: Gross domestic product per capita in US dollars

**lifeExpF**: female life expectancy in years

**pctUrban**: urban population in %

**infantMortality**: Infant mortality at the age of 1 year per 1000 live births

**groups**: OECD, Africa, Other

**region**: Africa, Asia, Caribbean, Europe, Latin Amer, North America, NorthAtlantic, Oceania


## Loading/cleaning the data and descriptive statistics

```{r libraries, echo=FALSE, warning= FALSE, message=FALSE}

library(tidyverse)
library(dplyr)
library(ggpubr)
library(car)
library(rstatix)
library(datasets)
library(dunn.test)
library(effsize)
library(plotly)

```


```{r datset, warning= FALSE}
data(UN)
head(UN)

```

```{r NAs, warning= FALSE, results='hide'}
#omit NAs
anyNA(UN)
#TRUE
UN_data <- na.omit(UN)

```

```{r summary, warning= FALSE}

#summary
summary(UN_data)

```
**Scatter plots for a basic understanding of the data**
```{r scatterplots, warning= FALSE, message = FALSE}

pairs(UN_data[, c("fertility", "ppgdp", "lifeExpF", "pctUrban", "infantMortality")],
      lower.panel = NULL) 

```


## Hypothesis
### H1 and H2

**H1**: There are significant differences between Africa and other country groups in terms of life expectancy and fertility rate.

**H2**: Life expectancy of women in African countries is significantly lower than in other country groups and in OECD countries. 


#### mean and sd by group

```{r mean overview group, warning= FALSE}

mean_overview_group <- UN_data %>%
  select(group, fertility, ppgdp, lifeExpF, pctUrban, infantMortality) %>%
  group_by(group) %>%
  mutate(N = n()) %>%  
  summarise_all(mean)

print(mean_overview_group)

```

```{r sd overview group, warning= FALSE}

sd_overview_group <- UN_data %>%
  select(group, fertility, ppgdp, lifeExpF, pctUrban, infantMortality) %>%
  group_by(group) %>%
  summarise_all(sd)

print(sd_overview_group)
```


#### Visualizations
Examinnation of the relationships between the variables using visualizations.

```{r histograms, warning= FALSE, message = FALSE}

hist(UN_data$lifeExpF, main = "", xlab = "Life expectancy of women in years")
hist(UN_data$fertility, main = "", xlab = "Fertility rate, number of children per woman")

```

```{r boxplots life exp fertility, warning= FALSE, message = FALSE}

ggplot(UN_data, aes(x = group, y = lifeExpF, color = group)) +
  geom_boxplot() +
  labs(
    x = "Group",
    y = "Life Expectancy of women (years)"
  )+
   scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))+
  theme_minimal()+
  guides(color = FALSE)


ggplot(UN_data, aes(x = group, y = fertility, color = group)) +
  geom_boxplot() +
  labs(
    x = "Group",
    y = "Fertility rate (number of children per woman)"
  ) +
  scale_color_manual(values = c("#00AFBB", "#E7B800", "#FC4E07"))+
  theme_minimal()+
  guides(color = FALSE)

```

```{r bubble chart life exp fertility, warning= FALSE, message = FALSE}

color <- c("#00AFBB", "#E7B800", "#FC4E07")
fig <- plot_ly(UN_data, x = ~fertility, y = ~lifeExpF,
               size = ~fertility, sizes = c(10, 1000), color = ~group, colors = color)
fig <- fig %>% layout(xaxis = list(title = "Fertility"),
                      yaxis = list(title = "Life expectancy"))
fig

```



#### Hypothesis testing

Checking the prerequisites for an ANOVA

- Normal distribution of the residuals

- Homogeneity of variance (Homoscedasticity)

**Testing the residuals for normal distribution**

The QQ plot compares the empirical quantiles of the residuals with the theoretical quantiles of a normal distribution. 
A good fit is indicated by an approximately straight-line relationship between the empirical and theoretical quantiles.

```{r qqplot fertility, warning= FALSE, message = FALSE}

#linear mode
model_fertility <- lm(fertility ~ group, data = UN_data)
#Q-Q-Plot
ggqqplot(residuals(model_fertility))

```

```{r qqplot life expectancy, warning= FALSE, message = FALSE}

#model
model_lifeExpF <- lm(lifeExpF ~ group, data = UN_data)
#Q-Q-Plot
ggqqplot(residuals(model_lifeExpF))

```

The residuals of both variables do not seem to be clearly normally distributed.
As a parametric method, one-way ANOVA provides the best interpretable results if the dependent variable is approximately normally distributed in each group. 

**Testing for homoscedasticity**

```{r levene test, warning= FALSE, message = FALSE, results = FALSE}

leveneTest(data= UN_data, lifeExpF ~ group)

# p=1.516e-06 < 0.05 there is no homogeneity of variance

leveneTest(data= UN_data, fertility ~ group)

# p= 4.358e-06 <0.05 there is no homogeneity of variance

```
Variance homogeneity is an important prerequisite for ANOVA, as it ensures that the variance of the dependent variables are approximately the same across the groups. If this assumption is violated, the ANOVA results may be unreliable and lead to erroneous conclusions.


```{r diagram, warning= FALSE, message = FALSE}

plot(model_fertility, 1)

plot(model_lifeExpF, 1)

```

The Residuals vs. Fitted diagram shows the residuals against the predicted values. A random pattern without systematic trends indicates homoscedasticity. The data points in the Residuals vs. Fitted Diagram in both cases almost form a triangle shape, it indicates heteroscedasticity, which means that the variance of the residuals is not constant across the range of fitted values.

Since the requirements for a one-way ANOVA are not sufficiently met, the hypothesis will be tested with the non-parametric Kruskal-Wallis-Test instead.

**Kruskal-Wallis-Test**

```{r kruskal wallis fertility, warning= FALSE, message = FALSE}

kruskal.test(fertility ~ group, data = UN_data)

```

```{r kruskal wallis life expectancy, warning= FALSE, message = FALSE}

kruskal.test(lifeExpF ~ group, data = UN_data)

```

The test result includes the chi-square value, the degree of freedom and the p-value. A significant p-value indicates that at least one group differs significantly from the other groups.

- Based on p-value < 2.2e-16 <0.05, it can be concluded that there is a statistically significant difference 
between the birth rates by country group.

- Based on p-value < 2.2e-16 <0.05, it can be concluded that there is a statistically significant difference 
between life expectancy by country group.

**post-hoc pairwise comparisons**

```{r dunns test fertility, warning= FALSE, message = FALSE}

dunn.test(UN_data$fertility, UN_data$group, method = "bonferroni")

```
```{r dunns test life exp, warning= FALSE, message = FALSE}

dunn.test(UN_data$lifeExpF, UN_data$group, method = "bonferroni")

```

If the p-value for a pairwise comparison is smaller than the significance level divided by the number of comparisons (alpha/2 in this case), then the null hypothesis will be rejected. The null hypothesis states that there is no difference between the groups being compared. The Bonferroni correction is used to reduce the probability of 
wrong results.
All pairwise comparisons indicate significant differences between African countries and OECD countries and others in terms of fertility rate and life expectancy. 

### H3

**H3**: There are significant differences between the regions Asia and Europe in terms of the urbanisation rates.

#### mean and sd by region

```{r mean overview region, warning= FALSE}

mean_overview_region <- UN_data %>%
  select(region, pctUrban) %>%
  group_by(region) %>%
  mutate(N = n()) %>% 
  summarise_all(mean)

print(mean_overview_region)
```

```{r sd overview region, warning= FALSE}

sd_overview_region <- UN_data %>%
  select(region, pctUrban) %>%
  group_by(region) %>%
  summarise_all(sd)

print(sd_overview_region)
```
#### Visualization
Examinnation of the relationships between the variables using a boxplot.

```{r boxplots urban, warning= FALSE, message = FALSE}

ggplot(UN_data, aes(x = region, y = pctUrban, color = region)) +
  geom_boxplot() +
  labs(
    x = "Region",
    y = "Urbanisation rate in %"
  )+
   scale_color_manual(values = c("#c0c0c0", "#FC4E07", "#c0c0c0", "#E7B800", "#c0c0c0", "#c0c0c0", "#c0c0c0"))+
  theme_minimal()+
  guides(color = FALSE)

```

#### Hypothesis testing

Checking the prerequisites for an ANOVA

- Normal distribution of the residuals

- Homogeneity of variance (Homoscedasticity)

**Testing the residuals for normal distribution**

```{r qqplot urban, warning= FALSE, message = FALSE}

model_urban <- aov(pctUrban ~ region, data = UN_data)

#Q-Q-Plot
ggqqplot(resid(model_urban))


```

```{r shapiro, warning= FALSE, message = FALSE, results= FALSE}

#shapiro-Wilk-Test
residuals_model <- shapiro.test(resid(model_urban))
print(residuals_model)
#p-value = 0.2188 >0.05 indicates normal distribution

```

The residuals of the dependent variable seems to be approximately normally distributed.

**Testing for Homoscedasticity**

```{r levene test urban, warning= FALSE, message = FALSE, results = FALSE}

leveneTest(data = UN_data, pctUrban ~ region)

# p=1.378e-05 < 0.05 Homogeneity of variance is not assumed

```

```{r diagram urban, warning= FALSE, message = FALSE}

plot(model_urban, 1)

```

Variance homogeneity can be assumed (even though the levene test had a different result) because the data points are relatively evenly distributed.
Since the requirements for a one-way ANOVA, normal distribution of the residuals and variance homogeneity are met, 
the one-way ANOVA will be used to test the hypothesis.

**one-way ANOVA**

```{r anova, warning= FALSE, message = FALSE}

anova(model_urban)

```

The p-value 1.222e-08 indicates that there are significant differences between the regions, which means that at least one of the regions differs statistically significantly from the others.

**post-hoc pairwise comparisons**

```{r pairwise t test, warning= FALSE}

UN_data %>%
  pairwise_t_test(pctUrban~region,
                  pool.sd = TRUE,
                  p.adjust.method = "bonferroni") %>%
  as.data.frame()

```

**Calculating the effect size**
How big are the differences between Africa and Asia in terms of the urbanisation rate?

```{r effect size, warning= FALSE}

subset_data <- subset(UN_data, region %in% c("Africa", "Asia"))

#cohens d comparison between Africa and Asia
cohen.d(pctUrban ~ region, data = subset_data)


```

Cohen's d measures the standard deviation of the differences between the groups in relation to the mean standard deviation. A value of -0.6927299 indicates that there is a medium effect.
